Spatial Reasoning and Statistical Graphics

### 2014-10-11 Paper/Draft1/TestingVisualAptitude.Rnw
first responses

\section{Introduction}
Relevant literature: 
\begin{itemize}
\item \citet{shah1995conceptual} showed that spatial ability was not correlated with accuracy on a simple two-dimensional line graph description task, but that mathematical ability was correlated with accuracy. 
\item \citet{just1985cognitive} showed that high-spatial-ability viewers used different rotation strategies than low-spatial-ability viewers when asked to whether three-dimensional alphabet cubes were the same. 
\item \citet{hofmann2012graphical} for lineup stimuli and general lineup performance
\end{itemize}

@@

\section{Results}
\newtext{Results are based on an evaluation of \Sexpr{nrow(ans)} undergraduate students at Iowa State University. }
Scoring of all test results was done such that random guessing leads to an expected value of 0; thus for a test consisting of multiple choice questions with $k$ suggested answers with a single correct answer each, the score is calculated as
\begin{eqnarray}\label{eq.scoring}
\# \text{total correct answers } - 1/(k-1) \cdot \# \text{wrong answers}.
\end{eqnarray}
This allows us to compare each participant's score in light of how many problems were attempted as well as the number of correct responses. Combining accuracy and speed into a single number does not only make a comparison of test scores easier,  this scoring mechanism is also used on many standardized tests, such as the SAT and the battery of psychological tests which parts of this test are drawn from \citep{diamond1973correction, ekstrom1976manual}.

@@

\comment{\textit{We might need to expand this a bit to account for the figure classification section, which has some questions with 2 answers and some with 3. Or we could leave it as an exercise for the reader :)}}

\comment{\textit{So are we scaling things by the variance/std dev. under random guessing? Or by the theoretical test range?}\\
Right now, I'm assuming that we're scaling between 0 and 100 by test range $[-a, b]$. So we take the guessing-fixed score and add $a$, then divide by $(a+b)$ and multiply by 100.}

@@

\comment{What happened to the arts skills? I seem to have a positive association now - why is my data different from yours? \\
\textit{I have the same updated results you do. I see that Stephanie changed the data file (but didn't add any more rows)... is it possible something changed there? Otherwise, maybe my dropbox didn't update correctly the first time? Or maybe it was figure caching?}}
Results are presented graphically in figures \ref{fig:visualaptitudects} and \ref{fig:visualaptitudecat}.

<<VisReasoningAnalysis,echo=FALSE, include=FALSE, fig.width=10,fig.height=5, out.width='\\linewidth'>>=
lineup.summary$variable <- factor(lineup.summary$variable, labels=c("Card Rotation", "Figure Classification", "Paper Folding", "Visual Search"))

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot") + facet_wrap(~variable, scales="free_x") + xlab("") + ylab("Lineup Score (# Correct - 1/19 # Incorrect)") + theme_bw() + geom_point(aes(x=value, y=lineup), shape=1, size=3)
ggsave(paste(imgdir, "fig-VisReasoningAnalysisCat.png", sep=""), width=8, height=6)
@

@@

\begin{figure}[htbp]\centering
\includegraphics[width=\textwidth]{fig-VisReasoningAnalysisCts}
\caption[Visual Aptitude Study Results]{Preliminary results of continuous variables compared to lineup score.}\label{fig:visualaptitudects}
\end{figure}

>>>>

\comment{The text makes still a lot of references to `preliminary results'. Don't just use copy \& paste for the changes!!!! Go through those sections and adjust. We are working on the final analysis now. Even if numbers might still change, they are not preliminary anymore. }
\section{Introduction}
Relevant literature: 
\begin{itemize}
\item \citet{shah1995conceptual} showed that spatial ability was not correlated with accuracy on a simple two-dimensional line graph description task, but that mathematical ability was correlated with accuracy. 
\item \citet{just1985cognitive} showed that high-spatial-ability viewers used different rotation strategies than low-spatial-ability viewers when asked to whether three-dimensional alphabet cubes were the same. 
\item \citet{hofmann2012graphical} for lineup stimuli and general lineup performance
\end{itemize}

@@

\section{Results}
\newtext{Results are based on an evaluation of \Sexpr{nrow(ans)} undergraduate students at Iowa State University. }
Scoring of all test results was done such that random guessing leads to an expected value of 0; thus for a test consisting of multiple choice questions with $k$ suggested answers with a single correct answer each, the score is calculated as
\begin{eqnarray}\label{eq.scoring}
\# \text{correct answers } - 1/(k-1) \cdot \# \text{wrong answers}.
\end{eqnarray}
This allows us to compare each participant's score in light of how many problems were attempted as well as the number of correct responses. Combining accuracy and speed into a single number does not only make a comparison of test scores easier,  this scoring mechanism is also used on many standardized tests, such as the SAT and the battery of psychological tests \citep{diamond1973correction, ekstrom1976manual} which parts of this test are drawn from.

@@

\comment{\textit{We might need to expand this a bit to account for the figure classification section, which has some questions with 2 answers and some with 3. Or we could leave it as an exercise for the reader :)}
It's fairly trivial to apply this to the mixture - the only assumption that goes into the variance discussion is that the questions are independent. We don't need to include it here, but you should apply it to the figure classification set.

}

\comment{\textit{So are we scaling things by the variance/std dev. under random guessing? Or by the theoretical test range?}\\
Let's use the standard deviation, so that all tests have an expected value of 0 and a variance/standard deviation of 1 under random guessing.}

@@

% \comment{What happened to the arts skills? I seem to have a positive association now - why is my data different from yours? \\
% \textit{I have the same updated results you do. I see that Stephanie changed the data file (but didn't add any more rows)... is it possible something changed there? Otherwise, maybe my dropbox didn't update correctly the first time? Or maybe it was figure caching?}}
Results are presented graphically in figures \ref{fig:visualaptitudects} and \ref{fig:visualaptitudecat}.

<<VisReasoningAnalysis,echo=FALSE, include=FALSE, fig.width=8,fig.height=8, out.width='0.9\\textwidth'>>=
lineup.summary$variable <- factor(lineup.summary$variable, labels=c("Card Rotation", "Figure Classification", "Paper Folding", "Visual Search"))

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot") + facet_wrap(~variable, scales="free_x") + xlab("Value") + ylab("Lineup Score (# Correct - 1/19 # Incorrect)") + theme_bw() + geom_point(aes(x=value, y=lineup), shape=1, size=3)
ggsave(paste(imgdir, "fig-VisReasoningAnalysisCat.png", sep=""), width=6, height=6)
@

@@

\begin{figure}[htbp]\centering
\includegraphics[width=0.9\textwidth]{fig-VisReasoningAnalysisCts}
\caption[Visual Aptitude Study Results]{Scatterplots of all test scores compared to participants' scores in the lineup tests. \comment{What is the main message of these plots? Relatively strong positive correlation to all established test scores?}}\label{fig:visualaptitudects}
\end{figure}
